---
title: "Airbnb Listing Price Prediction Analysis"
author: "Yash Singh"
output: word_document
---

------------------------------------------------------------------------

# ***Introduction***

This report presents the analysis and prediction of Airbnb listing prices using R. We'll cover data import, cleaning, exploratory data analysis (EDA), modeling, evaluation, and findings.

------------------------------------------------------------------------

# *Importing data and required library*

```{r}
library(readr)
library(dplyr)
library(ggplot2)

listings <- read_csv("/Users/yashsingh/Desktop/listings.csv")
```

The dataset used in this analysis is from Airbnb's official website, containing essential information such as listing id, price, host details, and geographical data.

# *Cleaning data and transformation*

```{r}
# Check for missing values
missing_values <- colSums(is.na(listings))
print(missing_values)

# Clean data: Drop columns with high missing values, remove rows with missing price,
# and fill missing reviews_per_month
listings_cleaned <- listings %>%
  select(-neighbourhood_group, -license) %>%
  filter(!is.na(price)) %>%
  mutate(reviews_per_month = ifelse(is.na(reviews_per_month), 0, reviews_per_month))


listings_cleaned <- listings_cleaned %>%filter(!is.na(last_review))

```

1.  **Handling Missing Values**: We initially checked for missing values across the dataset and decided to drop columns with high missing values (`neighbourhood_group`, `license`), remove rows with missing price values, and filled in missing `reviews_per_month` values with zeros.

# *Exploratory Data Analysis (EDA)*

## 1. Plot histogram of price distribution

```{r}
  ggplot(listings_cleaned, aes(x = price)) +
  geom_histogram(binwidth = 100, fill = "skyblue", color = "black") +
  labs(x = "Price", y = "Frequency", title = "Histogram of Listing Prices")+
    scale_x_continuous(breaks = seq(0, 1500, by = 500))

```

**Outcome :** Most of the listings on AirBnB our priced under 500 dollars

## 2. Scatter plot of reviews vs. availability

```{r}
ggplot(listings_cleaned, aes(x = number_of_reviews, y = availability_365)) +
  geom_point(alpha = 0.7, color = "blue") +
  labs(x = "Number of Reviews", y = "Availability (in days)", title = "Impact of Reviews on Availability")


```

**Outcome** : More number of reviews results in less availability which  indicates that reviews significantly increases chances of getting more customers.

## 3. Bar plot of room types

```{r}
ggplot(listings_cleaned, aes(x = room_type, fill = room_type)) +
  geom_bar() +
  labs(x = "Room Type", y = "Count", fill = "Room Type", title = "Distribution of Room Types")

```

The bar plot illustrates the distribution of room types among Airbnb listings, with entire homes/apartments being the most common, followed by private rooms and shared rooms.

## 4.Barplot of avg prices by neighborhood

```{r}
neighbourhood_summary <- listings_cleaned %>%
  group_by(neighbourhood) %>%
  summarize(total_price = mean(price))

# Plot total price by neighborhood
ggplot(neighbourhood_summary, aes(x = neighbourhood, y = total_price)) +
  geom_col(fill = "skyblue", color = "black") +
  labs(x = "Neighborhood", y = "avg Price", title = "Total Price of Listings by Neighborhood") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# *Adding days since last review as a feature*

```{r}
listings_cleaned <- listings_cleaned %>%
  mutate(days_since_last_review = as.numeric(Sys.Date() - last_review))
```

A new column `days_since_last_review` which tells us the number of days since they received a review. Listings with recent reviews may indicate that the property is well-maintained and actively managed.

# Modelling

## *1. Split data into training and testing sets*

```{r}
#remove all non numerical features first
listings_cleaned <- listings_cleaned %>%
  select(price,number_of_reviews,reviews_per_month,availability_365,days_since_last_review)

# normalize the data
normalize <- function(x){
  a<-(x-min(x))/(max(x)-min(x))
  return(a)
}
listings_cleaned<- listings_cleaned %>% mutate_all(normalize)

set.seed(1234)  # Set seed for reproducibility
train_index <- sample(nrow(listings_cleaned), 0.7 * nrow(listings_cleaned))
train_data <- listings_cleaned[train_index, ]
test_data <- listings_cleaned[-train_index, ]
```

-   First, we remove all non-numerical data from the `listining_cleaned` data set.

-   Now, we set the seed so that for this particular scenario we only randomize the data once rather than randomizing it every time we run the code.

-   We randomly split the data-set into training (70%) and testing (30%) sets to develop and evaluate our predictive model.

## *2.Build linear regression model*

```{r}
model_lm <- lm(price ~ days_since_last_review, data = train_data)

```

-   Now we make a linear regression model using the lm(linear model) function.

-   price in this scenario is the dependent variable.Change in independent variable will result in change of dependent variable.

-   In this case we have the independent variable as all the features in our data set.

# Model Evaluation

## 1. Predict using the linear regression model

```{r}
predictions_lm <- predict(model_lm, newdata = test_data)
```

## 2. Count how many predictions are within the margin

```{r}
correct_predictions <- sum(abs(predictions_lm - test_data$price))
```

## 3. Calculate total valid predictions and accuracy

```{r}
total_predictions <- length(predictions_lm)
accuracy <- correct_predictions / total_predictions 
cat("Accuracy of Prediction:", accuracy * 100, "%\n")

#linear model graph
ggplot(listings_cleaned,aes(x=price,y=days_since_last_review))+
  geom_point(alpha=0.7)+
  stat_smooth(method = "lm",col="red")
```

## 4. Calculate RMSE to check error margin

```{r}
rmse <- sqrt(mean((test_data$price - predictions_lm)^2))
print(paste("RMSE:", round(rmse, 2)))
```

# Conclusion

In conclusion, our analysis highlights the predictive capabilities of the linear model in predicting Airbnb listing prices. Through thorough data cleaning, exploratory analysis, and model evaluation, we've provided vital insights in pricing that can benefit for both hosts and future guests.
